Effectful Semantics for Substructural SSA

### 2023-11-29 pldi24/paper.tex
Started work on resource algebra
Removed warnings

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation}
  \begin{small}
\begin{BVerbatim}[baseline=c]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
  \end{small}
\qquad \qquad \centernot\implies \qquad \qquad
  \begin{small}
\begin{BVerbatim}[baseline=c]
let y = print("world");
let z = print("hello");
...
\end{BVerbatim}
  \end{small}
\end{equation}
In this case, substituting a term for a variable changes the order of I/O
actions, which changes the program's observable behaviour. On the other hand,
suppose we have an effect which increments a counter. Then in the following
program the same transformation is sound, since \(x\) is used exactly once:
\begin{equation}
  \begin{small}
\begin{BVerbatim}[baseline=c]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
  \end{small}
\qquad \qquad \implies \qquad \qquad
  \begin{small}
\begin{BVerbatim}[baseline=c]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
  \end{small}
\end{equation}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
  \begin{small}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
return 5 
\end{BVerbatim}
  \end{small}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{small}
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{small}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
  \begin{small}
\begin{BVerbatim}[baseline=c]
let x = nondet();
return 5 
\end{BVerbatim}
  \end{small}
\qquad \qquad \implies \qquad \qquad
  \begin{small}
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
  \end{small}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
  \begin{small}
\begin{BVerbatim}[baseline=c]
let x = nondet();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
  \end{small}
\qquad \qquad \centernot\implies \qquad \qquad
  \begin{small}
\begin{BVerbatim}[baseline=c]
let y = nondet();
let z = nondet();
return y == z
\end{BVerbatim}
  \end{small}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{nondet()}, and
can change this from a program which always returns \texttt{true} to one which
can return either \texttt{true} or \texttt{false}. On the other hand, if the
effect is the possibility of nontermination, then the substitution is sound.
\begin{equation}
\begin{small}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\end{small}
\qquad \qquad \implies \qquad \qquad
\begin{small}
\begin{BVerbatim}[baseline=c]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{small}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure term as one which is all three of central,
copyable, and discardable.

In addition to working with effects, compiler intermediate representations also
have to track accesses to memory. This information is typically gathered via
ad-hoc alias analyses, but a compositional alternative to these analyses is to
track this information via substructural type systems such as linear and affine
types, which associate aliasing and ownership information with individual
values. Having aliasing information justifies important optimisations such as
loop hoisting: 
\begin{equation}
  \begin{small}
\begin{BVerbatim}[baseline=c]
while (e) {
  x = load(a);
  body
}
\end{BVerbatim}
\end{small}
\qquad \qquad \implies \qquad \qquad
\begin{small}
\begin{BVerbatim}[baseline=c]
x = load(a);
while(e) {
  body
}
\end{BVerbatim}
\end{small}
\label{eqn:hoist}
\end{equation}
This transformation is only sound if the body of the loop \texttt{body} does not
access \texttt{a}, which is precisely the kind of frame condition that linear
types can track. Moreover, since we already want to track whether effectful
\emph{expressions} can be duplicated or discarded, the very same machinery can
be reused to track whether \emph{values} are linear/affine/relevant.

@@

In general, we often want to justify optimizations which \textit{refine} a
program by, for example, reducing the set of allowable behaviours, rather than
simply rewrite the program to an equivalent one.

>>>>

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation*}
\begin{BVerbatim}[baseline=c,fontsize=\small]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
let y = print("world");
let z = print("hello");
...
\end{BVerbatim}
\end{equation*}
In this case, substituting a term for a variable changes the order of I/O
actions, which changes the program's observable behaviour. On the other hand,
suppose we have an effect which increments a counter. Then in the following
program the same transformation is sound, since \(x\) is used exactly once:
\begin{equation*}
\begin{BVerbatim}[baseline=c,fontsize=\small]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
\end{equation*}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
\begin{BVerbatim}[baseline=c,fontsize=\small]
let x = may_loop();
return 5 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
return 5
\end{BVerbatim}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
\begin{BVerbatim}[baseline=c,fontsize=\small]
let x = nondet();
return 5 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
return 5
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
\begin{BVerbatim}[baseline=c,fontsize=\small]
let x = nondet();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
let y = nondet();
let z = nondet();
return y == z
\end{BVerbatim}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{nondet()}, and
can change this from a program which always returns \texttt{true} to one which
can return either \texttt{true} or \texttt{false}. On the other hand, if the
effect is the possibility of nontermination, then the substitution is sound.
\begin{equation}
\begin{BVerbatim}[baseline=c,fontsize=\small]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure term as one which is all three of central,
copyable, and discardable.

In addition to working with effects, compiler intermediate representations also
have to track accesses to memory. This information is typically gathered via
ad-hoc alias analyses, but a compositional alternative to these analyses is to
track this information via substructural type systems such as linear and affine
types, which associate aliasing and ownership information with individual
values. Having aliasing information justifies important optimisations such as
loop hoisting: 
\begin{equation}
\begin{BVerbatim}[baseline=c,fontsize=\small]
while (e) {
  x = load(a);
  body
}
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c,fontsize=\small]
x = load(a);
while(e) {
  body
}
\end{BVerbatim}
\label{eqn:hoist}
\end{equation}
This transformation is only sound if the body of the loop \texttt{body} does not
access \texttt{a}, which is precisely the kind of frame condition that linear
types can track. Moreover, since we already want to track whether effectful
\emph{expressions} can be duplicated or discarded, the very same machinery can
be reused to track whether \emph{values} are linear/affine/relevant.

@@

In general, we often want to justify optimizations which \textit{refine} a
program by, for example, reducing the set of allowable behaviours, rather than
simply rewrite the program to an equivalent one. One place where refinement
pops up quite often is in the treatment of undefined values and undefined 
behaviour; for example, in a language where reading a dangling pointer is UB,
